How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-01-08 doc/methods.tex
condensed sections into paper.tex -- lots of references added and in bib file

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species CITATIONS. However, this practice is common with marine microfossil data CITATIONS. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from Cramer et al. CITATION (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}

>>>>

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction (discussed below).

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. Because the estimated ages of each occurrence is a product of core-specific age-models and can be overly precise, the hope is that by binning the data this smooths over the between-core heterogeneity and thus homogenizes our disparate data sources. The occurrence histories of each species were then given binary codes used to model the presence or extinction of those species. For every occurrence of a species, except the last, that species is considered to have survived and was marked with a 0. The last occurrence of that species is considered the bin in which the taxon has gone extinct -- and is assigned a 1. This protocol means that we are reading the fossil record ``as written,'' a practice that is potentially dangerous as it is a overconfident statement of preservation and may be shortening the actual duration of that species \citep{Alroy2010,Alroy2000b,Alroy2014,Foote1997,Foote1999a,Foote2001,Foote1996e,Lloyd2012b,Marshall1995,Wang1996}. However, this practice is common with marine microfossil data due to their exceptional preservation rate \citep{Ezard2013,Ezard2016,Ezard2011,Liow2010}. In fact, with marine microfossils collected from cores a bigger problem may be over extending the duration of a species due to mixing and smearing within the cores CITATIONS.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from \citet{Cramer2011} (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}
